Attrition and Achievement Gaps in Online Learning

### 2015-01-25 paper_v3.Rnw
final version

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments. A systematic analysis of individual differences in attrition and grades in 20 massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap. Online learners in Africa, Asia, and Latin America scored substantially lower grades and were up to half as likely to persist than those in Europe, Oceania, and Northern America. Women also exhibited lower persistence and performance then men. Yet more persistent learners were only marginally more satisfied with their achievement. The primary obstacle for most learners was finding time for the course, which was partly related to low levels of volitional control. Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones ($N=756$). 
Insights into why learners leave online courses informs models of attrition and targeted interventions to support learners achieve their goals.
\end{abstract}

@@

Educational environments have become increasingly diverse in format. Traditional schools and universities have a characteristically rigid structure, including instructor-defined syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs resemble those on online media platforms \cite{kizilcec2013deconstructing,breslow2013studying}. This trend is also reflected in the large spectrum of learners' motivations for enrolling in MOOCs \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Perspectives on attrition in open online courses ultimately depend on how these courses are viewed. For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}. This view of MOOC participation clearly adds complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total. Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals. These learners could benefit from changes to the course or targeted interventions. We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition. Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance. For Study 2, we invited around 6,053 learners who were predicted to have disengaged from a MOOC to provide feedback. We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time. Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting. The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno1993role}. Motivation alone is necessary but not sufficient for persistence, because students may fail to sustain efforts in the absence of strong self-regulatory skills.

@@

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments (e.g. \cite{walton2007question}).

\begin{description}
  \item[RQ1] How are prior characteristics (demographics, geographic location, intentions, and prior experience) related to learners' likelihood of disengaging from the course?
  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
\end{description}

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}. Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (instilling a growth mindset \cite{dweck2006mindset}) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}. This highlights the critical role of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.

\begin{description}
  \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
  \item[RQ2b] How does learners' level of satisfaction depend on their behavior in the course?
  \item[H3] Successful learners have a more growth mindset than unsuccessful learners.
\end{description}

@@

Distance education, relative to traditional in-person education, tends to attract a more mature student demographic and provide fewer opportunities for social integration. Students in distance learning programs are more likely to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.). The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

@@

\section{Study 1: Attrition \& Satisfaction}

The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners. This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.

@@

Twenty MOOCs on topics in a variety of disciplines were selected for investigation. Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment. On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf. Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region. Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.

@@

A number of significant and substantial individual differences emerged. Women were 12 to 20\% less likely than men to persist with lectures and assessments. Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.  Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$). In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America. While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America. Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe. Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners. More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.

The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work presents evidence for the existence of these gaps, but cannot explain why they occur. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. A geographical categorization to compare countries with a high and low propensity of English speakers could be misleading, due to confounding with psychological barriers that are less likely to hinder learners in English-speaking countries. More work is needed to identify potential causal mechanisms underlying these gaps.

@@

\subsubsection{Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

Attrition can be a sign of success if a learner has achieved their personal learning objectives. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$). Satisfaction ratings were compared using mixed-effects regression ($N=23,390$). Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$). As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential. Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them. The following analysis of reasons for attrition can shed more light on this hypothesis.

\subsubsection{Reasons for Attrition}
% RQ3a: What reasons do learners report for disengaging from a MOOC?
% RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?

A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a pre-defined list ($RQ3a$). This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses. Figure \ref{fig:s1reason} illustrates how frequently each reason was selected across courses. Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating. In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gains. The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course. Alternatively, a low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. Learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control. We investigate this hypothesis in Study 2.

Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents. Notably, 17\% of respondents in a typical course stopped participating because they learned all they intended to learn. This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success. It also supports the observation that progress in the course and satisfaction were only weakly related. We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following four-factor solution emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines);  second, {\em difficulty} (too advanced, exam too difficult); third, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well, not what I'm looking for, technical difficulties), and fourth, {\em goals \& expectations} (only exploring, learned all I wanted, not challenging, credential not valuable). The `late start' item failed to load onto any factor. General time issues and difficulty are very interpretable factors, while the other two factors comprise a complex set of reasons that should probably be further separated.

@@

\section{Study 2: Understanding Attrition}

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We collected feedback from learners who were likely to have stopped following the course. They were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners. Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learner's volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.). In contrast to attrition due to important obligations, there exist techniques to help learners become more self-regulated, which could boost their persistence. We constructed the following hypotheses about the role of volition in time-related attrition:

@@

\begin{table*}[ht]
\caption{Descriptive and inferrential statistics for psychological measures by self-ascribed success}
\label{tab:psych}
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

We further investigated research question $RQ3a$ about reasons for disengaging from the course, and tested hypotheses $H4ab$ about volitional control. In a second set of analyses, we tested our hypotheses about the role of psychological factors in the self-ascribed success of MOOC learners.
\\
\subsubsection{Psychological Factors}

>>>>

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments. A systematic analysis of individual differences in attrition and performance in 20 massive open online courses ($N>67,000$) revealed a geographic achievement gap and a gender achievement gap. Online learners in Africa, Asia, and Latin America scored substantially lower grades and were only half as likely to persist than those in Europe, Oceania, and Northern America. Women also exhibited lower persistence and performance than men. Yet more persistent learners were only marginally more satisfied with their achievement. The primary obstacle for most learners was finding time for the course, which was partly related to low levels of volitional control. Self-ascribed successful learners reported higher levels of goal striving, growth mindset, and feelings of social belonging than unsuccessful ones. Insights into why learners leave online courses inform models of attrition and targeted interventions to support learners achieve their goals.
\end{abstract}

@@

Educational environments have become increasingly diverse in format. Traditional schools and universities have a characteristically rigid structure, including instructor-defined syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs have delivered education to remote parts of the world and to people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion would only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs resemble those on online media platforms \cite{kizilcec2013deconstructing,breslow2013studying}. This trend is also reflected in the large spectrum of learners' motivations for enrolling in MOOCs \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Perspectives on attrition in open online courses ultimately depend on how these courses are viewed. For instance, MOOCs can be viewed as bridging two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}. This view of MOOC participation clearly adds complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total. Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggling to achieve their learning goals. These learners could benefit from changes to the course or targeted interventions. We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition. Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance. For Study 2, we invited around 6,053 learners who were predicted to have disengaged from a MOOC to provide feedback. We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time. Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or drop out of a traditional higher education setting. The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course. An early model suggested that students' persistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno2001volitional}. Motivation alone is necessary but not sufficient for persistence, because students may fail to sustain efforts in the absence of strong self-regulatory skills.

@@

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of individual students' characteristics, prior experiences, and experiences during college. While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments (e.g. \cite{walton2007question}).

\begin{description}
  \item[RQ1] How are learners' characteristics (demographics, geographic location, intentions, prior experience) related to their persistence and performance in the course?
  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
\end{description}

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}. Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (instilling a growth mindset \cite{dweck2006mindset}) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}. This highlights the critical role of mindset and cautions against overinterpreting non-persistence as an indicator of goal achievement.

\begin{description}
  \item[RQ2] How is learners' satisfaction related to their persistence in the course?
  \item[H3] Successful learners have a stronger growth mindset than unsuccessful learners.
\end{description}

@@

Distance education, relative to traditional in-person education, tends to attract a more mature student demographic and provide fewer opportunities for social integration. Students in distance learning programs are more likely to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.). The significant change from Tinto's original model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

@@

\section{Study 1: Individual Differences, Satisfaction, \& Reasons for Attrition}

The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners. This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggling learners.

@@

Twenty MOOCs on topics in a variety of disciplines were selected for investigation. Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment. On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intended to spend on the course. On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf. Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region. Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.

@@

A number of significant and substantial individual differences emerged. Women were 12 to 20\% less likely than men to persist with lectures and assessments. Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.  Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$). In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessments than learners in Northern America. While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America. Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe. Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners. More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistence. Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.

Leveraging the large scale of MOOCs, this study has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work presents evidence for the existence of these gaps, but cannot explain why they occur. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. A geographical categorization to compare countries with a high and low propensity of English speakers could be misleading, due to confounding with psychological barriers that are less likely to hinder learners in English-speaking countries. More work is needed to identify potential causal mechanisms underlying these gaps.

@@

\subsubsection{Satisfaction}

Attrition can be a sign of success if a learner has achieved their personal learning objectives. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating and those who persisted to different extents ($RQ2$). Satisfaction ratings were compared using mixed-effects regression with multiple imputation of missing values ($N=23,390$). Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$). As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential. Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them. The following analysis of reasons for attrition can shed more light on this hypothesis.

\subsubsection{Reasons for Attrition}

A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a predefined list ($RQ3a$). This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses. Figure \ref{fig:s1reason} illustrates how frequently each reason was selected across courses. Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating. In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time. According to expectancy-value theory \cite{wigfield2000expectancy}, the learner is a rational agent who allocates time to different activities to maximize subjective gains. The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course. Alternatively, a low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. Learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control. We investigate this hypothesis in Study 2.

Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents. Notably, 17\% of respondents in a typical course stopped participating because they had learned all they intended to learn. This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success. It also supports the conclusion that progress in the course and satisfaction were only weakly related. We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following four-factor solution emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines);  second, {\em difficulty} (too advanced, exam too difficult); third, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well, not what I'm looking for, technical difficulties), and fourth, {\em goals \& expectations} (only exploring, learned all I wanted, not challenging, credential not valuable). The `late start' item failed to load onto any factor. General time issues and difficulty are very interpretable factors, while the other two factors comprise a complex set of reasons that should probably be further separated.

@@

\section{Study 2: Psychological Factors in Attrition}

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We collected feedback from learners who were likely to have stopped following the course. They were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners. Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal commitments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learner's volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.). In contrast to attrition due to important obligations, there exist techniques to help learners become more self-regulated, which could boost their persistence. We constructed the following hypotheses about the role of volition in time-related attrition:

@@

\begin{table*}[ht]
\caption{Descriptive and inferential statistics for psychological measures by self-ascribed success}
\label{tab:psych}
\small
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

First, we tested hypotheses about the role of psychological factors in the self-ascribed success of online learners. In a second set of analyses, we further investigated reasons for leaving a course ($RQ3a$) and the role of volition ($H4$).

\subsubsection{Psychological Factors}
